Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-06 paper.tex
Replace quotes

The findings of this study are based on sequences and metadata available on
GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
if they (1) had ambiguous collection dates; (2) were collected before January
01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
MN908947.3) using Nextclade v2.3.0 \citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, sequences were excluded if they had a “bad” quality control status
in any of the four Nextclade columns ("qc.missingData.status”,
“qc.mixedSites.status”, “qc.frameShifts.status”, and “qc.stopCodons.status”).
These excluded genomes are enriched for low-coverage, contaminated, or
misassembled sequences.

@@

Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise
genome alignments were encoded as N. Problematic bases in the alignments, which
had two or more ambiguous letters or gaps within a distance of seven bases,
were masked as missing data following the protocol in
\url{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored
when attaching a sample genome to the ARG, but they were subsequently imputed.
Further, 481 problematic sites flagged as prone to sequencing errors or highly
homoplasic were excluded entirely
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsubsection{Exclusion of “time travellers”}

Including samples with incorrectly reported collection dates (e.g., an Alpha
sample purportedly collected in 2020 from the United States, before Alpha
appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
artefacts that distorted the topology of ARGs inferred early on in the study.
Hence, to exclude such potential “time travellers”, we employed two filters.
The first was to remove any sequence with a collection date that pre-dated the
time to the most recent common ancestor (tMRCA) of its corresponding clade. The
tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
(downloaded on August 22, 2022) where we used the lower bound of the 95\%
confidence interval of each clade as the minimum date cut-off. The second
filter was a simple threshold on the time delay between the collection date and
submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
inferred from the MSD-filtered genome set had noticeably fewer branches with a
high number of reversions. After some preliminary analysis, a MSD threshold
value of 30 days was used when building the ARGs described herein. Future work
could estimate collection dates by adding these potential time travellers back
on the tree, allowing an automated assessment of collection date discrepancies.

@@

Each daily batch of genomes is first attached to the growing ARG without any
phylogenetic resolution. In cases where the newly added sequence is inferred to
be a recombinant, a recombination node is inserted; the “date” of this node is
arbitrarily set to the average of the collection dates of the genome and the
most recent of its parent nodes. Otherwise, the genomes are attached to the
most closely related ancestral node (an attachment node). This creates
polytomies that are then locally resolved among the descendants of each
attachment node using the Unweighted Pair Group Method with Arithmetic Mean
(UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
This hierarchical clustering algorithm produces a dendrogram relating a set of
sequences, which is treated as an ultrametric binary tree here. Subsequently,
mutations are placed onto the daily tree by maximum parsimony. An issue with
the UPGMA algorithm is that it may create internal nodes supported by no
informative site (i.e., having no mutation immediately ancestral to them). In
an adjustment step, such unsupported internal nodes are removed to “un-resolve”
the daily tree; the remaining internal nodes, which are supported by at least
one shared mutation, are retained. The “dates” of the non-sample internal nodes
are set to the same date as the samples in the daily batch. The resulting daily
tree is grafted onto the attachment node in the global ARG.

@@

To illustrate the use of sc2ts, we employed two different sampling strategies.
First, we inferred a densely sampled but time-limited “wide” ARG by attaching
all the sample genomes with an MSD of 30 days, one daily batch at a time. The
inference procedure ran until the genomes sampled on June 30, 2021 were
attached to the ARG, resulting in a total of 1.27 million samples. Second, we
inferred a “long” ARG over a more extended timespan, by attaching at most 1,000
genomes, which were randomly subsampled from each daily batch with an MSD
within 30 days. For this ARG, the inference procedure ran until the genomes
sampled on June 30, 2022 were attached to the ARG. The long ARG covers an extra
year of samples but is more sparsely sampled. June 30, 2022 is the collection
date of the final batch of sequences which were preprocessed for the
development of sc2ts. Once sc2ts is further developed and tested, we will
attach samples collected after that date to infer a fully sampled
pandemic-scale ARG.

@@

Although here we focus on a single local tree at an arbitrarily chosen
position, note that in the original “un-simplified” ARGs, there are 1,496 and
958 trees, respectively (Table 1). Hence, while for a relatively small sample
the genealogy may be well described by a single tree, the full dataset shows
substantial variation in phylogenetic relationships (or phylogenetic
incongruence) across the SARS-CoV-2 genome, much of which may be attributable
to recombination. Such complex reticulate relationships are appropriately
captured by a genealogy in the form of an ARG.

@@

For simplicity, we focus primarily on “non-complex” recombination nodes. To do
this, we apply a filtering step that removes those with more than one
breakpoint (and often more than two parent lineages), or where matching is not
HMM-consistent or identifies parents with different Pango lineage assignments
in the original (i.e., 5’ to 3’ end) and mirrored (i.e., 3’ to 5’ end) HMM
runs. Filtering reduces both the number of recombination nodes and breakpoints
from  A \& B to C in the wide ARG and from 726 \& 806 to 459 in the long ARG
(Supplementary Tables 1 and 2). Note that in the unfiltered case, a single
recombination node can correspond to multiple points, one for each
recombination breakpoint. The effect of this filtering step in the long ARG,
which includes time periods of increased rates of coinfection and recombination
compared to the wide ARG, can be seen in Figure 3A (unfiltered) and Figure 3B
(filtered).

@@

The mosaic genome structure of all the 32 Pango recombinants is concordant with
the structure determined by the Pango Lineage Designation Committee
(Supplementary Table 3). For example, the genome of XD (also known as
“Deltacron”) is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
(Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.

@@

TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006
in the UShER public phylogeny. Node 628656 (Scotland/QEUH-37794BB/2022) is a
sister clade of XAC, and annotated as “miscBA2BA1PostSpike”.

@@

“The rate of putative recombination breakpoints is about three times higher
towards the 3' of the change point than the 5' interval.“ Consistent with this,
the breakpoint intervals of the putative recombinants identified here tend to
occur on the 3’ end of the SARS-CoV-2 genome.

@@

\begin{itemize} \item A potential source of error in sc2ts ARGs comes from
incorrectly reported collection dates of the samples, which can cause sample
genomes to be attached to the wrong parts of the ARGs. Thus far, we have
attempted to correct the most offending errors in the ARGs by filtering out
"time travellers", but the sampling time can still be wrong by days or tens of
days. ***describe possible approaches to correct this?*** \item To avoid time
travellers, we used a maximum submission delay, but this will systematically
bias against submissions from countries with greater submission delays (Kalia
et al.). An area of future work is to add back the excluded sequences to a
built ARG to assess which ones are likely true time travellers. \item To avoid
time travellers, we used a maximum submission delay, but this will
systematically bias against submissions from countries with greater submission
delays (Kalia et al.). An area of future work is to add back the excluded
sequences to a built ARG to assess which ones are likely true time travellers.
\item We have used a single mismatch value in all the HMM runs, set so that
having three mutations is as likely as having a single recombination. This
parameter choice is arbitrary, and higher values may reduce the number of
origins for those Pango lineages that were inferred to descend from more than
one recombinant ancestor. In addition, because both the mutation rate and
recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
allow the mismatch value to vary along the genome. \item By treating mutations
equally in the current HMM implementation, we implicitly assume that the rates
of different types of mutations are identical. However, as reported previously
(Yi et al. 2021), there is a high degree of mutational asymmetry in SARS-CoV-2.
To account for mutational biases, more sophisticated nucleotide substitution
models (e.g., allowing for different rates of transition and transversion) can
be incorporated into HMM likelihood calculations. \item Our current masking
strategy handles indels as missing data. Doing this can remove information
helpful to detect recombinants. For example, sequences of the Pango recombinant
XP were not identified as recombinants, because XP’s characteristic multibase
deletion was masked. \item This "lack of memory" in the HMM can create
recombinants with three or more parents when it would be equally parsimonious
to have just two. The copying process in the HMM currently does not limit the
number of parents to two, but doing so would be a valuable addition. \item Some
recombinants may reasonably have three or more parents, indicating multiple
"stacked" recombination events. We have largely ignored these in the current
analysis, but with further sc2ts development, there is potential to analyse
these putative recombinants and trace their evolutionary origins through the
ARG. \item In the present implementation, HMM traceback is performed once,
yielding a single maximum-likelihood genome copying path. A stochastic
traceback algorithm (e.g., (Rasmussen et al. 2014)) can be implemented in order
to quantify uncertainty about putative parent node(s) and switch positions.
These estimates of uncertainty may be used as support values for branches
analogous to bootstrap values produced by standard phylogenetic methods. \item
Daily trees are built using the UPGMA algorithm. This entails an extra step to
“un-resolve” daily trees in parts of the trees with no phylogenetic resolution.
To avoid this step, a better approach is to use a maximum-parsimony algorithm
to build daily trees instead. \end{itemize}

>>>>

The findings of this study are based on sequences and metadata available on
GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
% FIXME
doi.org/10.55876/gis8.230329cd (Supplementary Table 1).
Sequences were removed
if they (1) had ambiguous collection dates;
(2) were collected before January 01, 2020 or after June 30, 2022; or
(3) were isolated from a non-human host.
Sequences were aligned to the
Wuhan-Hu-1/2019 reference sequence (GenBank: MN908947.3)
using Nextclade v2.3.0~\citep{Aksamentov2021-hj}
(dataset tag 2022-07-26T12:00:00Z).
To reduce the impact of poor-quality sequences on ARG inference,
sequences were excluded if they had a ``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and
``qc.stopCodons.status'').
These excluded genomes are enriched for low-coverage, contaminated, or
misassembled sequences.

@@

Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise
genome alignments were encoded as N. Problematic bases in the alignments, which
had two or more ambiguous letters or gaps within a distance of seven bases,
were masked as missing data following the protocol in
\url{https://github.com/roblanf/sarscov2phylo/}.
Masked bases were ignored when attaching a sample genome to the ARG,
but they were subsequently imputed.
Further, 481 problematic sites flagged as prone to sequencing errors or highly
homoplasic were excluded entirely
% Do we need this first URL?
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsubsection{Exclusion of ``time travellers''}

Including samples with incorrectly reported collection dates (e.g., an Alpha
sample purportedly collected in 2020 from the United States, before Alpha
appeared in the United Kingdom, USA/MN-Mayo-1563/2020) was found to produce
artefacts that distorted the topology of ARGs inferred early on in the study.
Hence, to exclude such potential ``time travellers'' we employed two filters.
The first was to remove any sequence with a collection date that pre-dated the
time to the most recent common ancestor (tMRCA) of its corresponding clade. The
tMRCA for each clade was obtained from a Nextstrain GISAID reference tree
(downloaded on August 22, 2022) where we used the lower bound of the 95\%
confidence interval of each clade as the minimum date cut-off. The second
filter was a simple threshold on the time delay between the collection date and
submission date (maximum submission delay, or MSD). Indeed, we found that ARGs
inferred from the MSD-filtered genome set had noticeably fewer branches with a
high number of reversions. After some preliminary analysis, a MSD threshold
value of 30 days was used when building the ARGs described herein. Future work
could estimate collection dates by adding these potential time travellers back
on the tree, allowing an automated assessment of collection date discrepancies.

@@

Each daily batch of genomes is first attached to the growing ARG without any
phylogenetic resolution. In cases where the newly added sequence is inferred to
be a recombinant, a recombination node is inserted; the ``date'' of this node is
arbitrarily set to the average of the collection dates of the genome and the
most recent of its parent nodes. Otherwise, the genomes are attached to the
most closely related ancestral node (an attachment node). This creates
polytomies that are then locally resolved among the descendants of each
attachment node using the Unweighted Pair Group Method with Arithmetic Mean
(UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
This hierarchical clustering algorithm produces a dendrogram relating a set of
sequences, which is treated as an ultrametric binary tree here. Subsequently,
mutations are placed onto the daily tree by maximum parsimony. An issue with
the UPGMA algorithm is that it may create internal nodes supported by no
informative site (i.e., having no mutation immediately ancestral to them). In
an adjustment step, such unsupported internal nodes are removed to ``un-resolve''
the daily tree; the remaining internal nodes, which are supported by at least
one shared mutation, are retained. The ``dates'' of the non-sample internal nodes
are set to the same date as the samples in the daily batch. The resulting daily
tree is grafted onto the attachment node in the global ARG.

@@

To illustrate the use of sc2ts, we employed two different sampling strategies.
First, we inferred a densely sampled but time-limited ``wide'' ARG by attaching
all the sample genomes with an MSD of 30 days, one daily batch at a time. The
inference procedure ran until the genomes sampled on June 30, 2021 were
attached to the ARG, resulting in a total of 1.27 million samples. Second, we
inferred a ``long'' ARG over a more extended timespan, by attaching at most 1,000
genomes, which were randomly subsampled from each daily batch with an MSD
within 30 days. For this ARG, the inference procedure ran until the genomes
sampled on June 30, 2022 were attached to the ARG. The long ARG covers an extra
year of samples but is more sparsely sampled. June 30, 2022 is the collection
date of the final batch of sequences which were preprocessed for the
development of sc2ts. Once sc2ts is further developed and tested, we will
attach samples collected after that date to infer a fully sampled
pandemic-scale ARG.

@@

Although here we focus on a single local tree at an arbitrarily chosen
position, note that in the original ``un-simplified'' ARGs, there are 1,496 and
958 trees, respectively (Table 1). Hence, while for a relatively small sample
the genealogy may be well described by a single tree, the full dataset shows
substantial variation in phylogenetic relationships (or phylogenetic
incongruence) across the SARS-CoV-2 genome, much of which may be attributable
to recombination. Such complex reticulate relationships are appropriately
captured by a genealogy in the form of an ARG.

@@

For simplicity, we focus primarily on ``non-complex'' recombination nodes. To do
this, we apply a filtering step that removes those with more than one
breakpoint (and often more than two parent lineages), or where matching is not
HMM-consistent or identifies parents with different Pango lineage assignments
in the original (i.e., 5’ to 3’ end) and mirrored (i.e., 3’ to 5’ end) HMM
runs. Filtering reduces both the number of recombination nodes and breakpoints
from  A \& B to C in the wide ARG and from 726 \& 806 to 459 in the long ARG
(Supplementary Tables 1 and 2). Note that in the unfiltered case, a single
recombination node can correspond to multiple points, one for each
recombination breakpoint. The effect of this filtering step in the long ARG,
which includes time periods of increased rates of coinfection and recombination
compared to the wide ARG, can be seen in Figure 3A (unfiltered) and Figure 3B
(filtered).

@@

The mosaic genome structure of all the 32 Pango recombinants is concordant with
the structure determined by the Pango Lineage Designation Committee
(Supplementary Table 3). For example, the genome of XD (also known as
``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
(Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.

@@

TODO: Node  740761 (USA/NC-CDC-LC0668306/2022) is a sister clade proposed1006
in the UShER public phylogeny. Node 628656 (Scotland/QEUH-37794BB/2022) is a
sister clade of XAC, and annotated as ``miscBA2BA1PostSpike''.

@@

``The rate of putative recombination breakpoints is about three times higher
towards the 3' of the change point than the 5' interval.`` Consistent with this,
the breakpoint intervals of the putative recombinants identified here tend to
occur on the 3’ end of the SARS-CoV-2 genome.

@@

\begin{itemize} \item A potential source of error in sc2ts ARGs comes from
incorrectly reported collection dates of the samples, which can cause sample
genomes to be attached to the wrong parts of the ARGs. Thus far, we have
attempted to correct the most offending errors in the ARGs by filtering out
"time travellers", but the sampling time can still be wrong by days or tens of
days. ***describe possible approaches to correct this?*** \item To avoid time
travellers, we used a maximum submission delay, but this will systematically
bias against submissions from countries with greater submission delays (Kalia
et al.). An area of future work is to add back the excluded sequences to a
built ARG to assess which ones are likely true time travellers. \item To avoid
time travellers, we used a maximum submission delay, but this will
systematically bias against submissions from countries with greater submission
delays (Kalia et al.). An area of future work is to add back the excluded
sequences to a built ARG to assess which ones are likely true time travellers.
\item We have used a single mismatch value in all the HMM runs, set so that
having three mutations is as likely as having a single recombination. This
parameter choice is arbitrary, and higher values may reduce the number of
origins for those Pango lineages that were inferred to descend from more than
one recombinant ancestor. In addition, because both the mutation rate and
recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
allow the mismatch value to vary along the genome. \item By treating mutations
equally in the current HMM implementation, we implicitly assume that the rates
of different types of mutations are identical. However, as reported previously
(Yi et al. 2021), there is a high degree of mutational asymmetry in SARS-CoV-2.
To account for mutational biases, more sophisticated nucleotide substitution
models (e.g., allowing for different rates of transition and transversion) can
be incorporated into HMM likelihood calculations. \item Our current masking
strategy handles indels as missing data. Doing this can remove information
helpful to detect recombinants. For example, sequences of the Pango recombinant
XP were not identified as recombinants, because XP’s characteristic multibase
deletion was masked. \item This "lack of memory" in the HMM can create
recombinants with three or more parents when it would be equally parsimonious
to have just two. The copying process in the HMM currently does not limit the
number of parents to two, but doing so would be a valuable addition. \item Some
recombinants may reasonably have three or more parents, indicating multiple
"stacked" recombination events. We have largely ignored these in the current
analysis, but with further sc2ts development, there is potential to analyse
these putative recombinants and trace their evolutionary origins through the
ARG. \item In the present implementation, HMM traceback is performed once,
yielding a single maximum-likelihood genome copying path. A stochastic
traceback algorithm (e.g., (Rasmussen et al. 2014)) can be implemented in order
to quantify uncertainty about putative parent node(s) and switch positions.
These estimates of uncertainty may be used as support values for branches
analogous to bootstrap values produced by standard phylogenetic methods. \item
Daily trees are built using the UPGMA algorithm. This entails an extra step to
``un-resolve'' daily trees in parts of the trees with no phylogenetic resolution.
To avoid this step, a better approach is to use a maximum-parsimony algorithm
to build daily trees instead. \end{itemize}
